Synthesizing Quotient Lenses

### 2018-03-10 papers/QuotientLenses/main.tex
in progress edits to intro; changed John Doe to Ada Lovelace

\section{Introduction}
Oftentimes, the same data is stored in a variety different formats. For
example, bibliography data may be stored in \bibtex{} and
EndNote formats; spreadsheet data can be represented as a CSV file
or a TSV file; web APIs can return data as a JSON or XML object. In such cases,
it is necessary to synchronize the two or more copies of data so that they
store the same information.

One way to synchronize data stored in different formats is to use \emph{lenses},
\emph{i.e.,} programs that represent pairs of functions for translating data
from format the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
derived using domain-specific bidirectional languages which guarantee that the
lenses defined in that language satisfy a certain set of \emph{lens laws} that
give guarantees on how the two transformations interact.

Amongst the most restrictive lenses are the \emph{bijective lenses}.
A bijective lens $\ell : S \Leftrightarrow T$ is simply a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since these lenses demand that get and put functions form a bijection, then
no information is lost when one translates from one data format to another. 
While this strong guarantee can be useful in some contexts, it is more often
the case that some of the information in one or both of the data formats is
superfluous. For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant. Many
data formats have other ad hoc rules as well. In bibliographies, for example,
recall that in \bibtex{}, an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not
matter which of the two formats is used.

{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions of
bijective string lenses by allowing a programmer to define an equivalence
relation on the source and target formats so that the round-tripping lens laws
hold modulo these equivalence rather than modulo equality. Quotient lenses thus
make it possible to align many more data sources than would otherwise be
possible, by ignoring details that a programmer views as inessential. The only
language we are aware of that implements a general class of quotient lenses is
{\em Boomerang}~\cite{boomerang}. While Boomerang enables a programmer to
define a very general class of quotient lenses, these lenses to have two main
drawbacks.

First, in order to define an equivalence relation in Boomerang, a programmer
must define a {\em canonizer}; pair of functions $f$ and $g$ where $f$ maps
data represented in the original external format to a canonical element and $g$
chooses which elements of the original format correspond to values in the
canonical format. The disadvantage of this approach is that in order for a
programmer to {\em specify} a quotient lens in Boomerang, she must define four
different data formats: two ``outer'' formats which specify the data that is to
be transformed by the lens, and two ``inner'' formats which specify the
canonical formats for the source and target data.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} or QREs, which enable a programmer to simultaneously specify a
regular expression as well as a set of canonical representatives for strings
that match the regular expression. QREs have the added advantage that once the
programmer has defined a well formed QRE, then she need not define a
canonization function seperately; the canonization function can automatically be
inferred from the QRE.

The second challenge with programming lenses in Boomerang is that even after the
programmer has specified the data as well as a canonizer that maps each member
in the source and target sets to their respective canonical representative, she
must still define the underlying lens that maps between the respective sets of
canonical representatives. This can be a slow and tedious task due to the
constraints that Boomerang imposes on its terms. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

Our solution to this problem is that we use QREs to define a class of quotient
lenses called {\em QRE lenses}, and, we show how to synthesize such QRE lenses
automatically. More specifically, in order to generate a lens, a programmer
provides a pair of QREs describing the source and target formats, plus
a (possibly empty) suite of examples that demonstrates how the
lens should act in specific cases.  Given this information, we can exploit and
extend recent work on algorithms for synthesis of bijective
lenses~\cite{optician} to generate quotient lenses. The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (which cuts down the amount of non-determinism in
a specification and greatly speeds up synthesis), we prove that
this rigidity does not compromise their expressiveness.

@@

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective String Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Doe,
Author = "John Doe",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Doe
%A John Doe
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

@@

\begin{center}
\begin{minipage}{2.5in}
\begin{lstlisting}
bib_example =
"@Book {Doe,
 Author = \"John Doe\",
 Title = {Generic Title},
}"
\end{lstlisting}
\end{minipage}
\begin{minipage}{1.5in}
\begin{lstlisting}
end_example =
"%0 Book
 %T Generic Title
 %A John Doe
 %F Doe"
\end{lstlisting}
\end{minipage}
\end{center}

>>>>

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made to data in one format, data in the other must be updated so the
two representations continue to store the same information.
One way to synchronize such data is to use \emph{lenses}\ksf{include citations?}, which
are programs that simultaneously define pairs of functions for translating
the data 
from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
defined using domain-specific bidirectional languages that 
ensure the specified lenses satisfy various \emph{lens laws} that
guarantee how the \emph{put} and \emph{get} functions interact.

\emph{Bijective lenses} are amongst the most restrictive. 
A bijective lens $\ell : S \Leftrightarrow T$ specifies a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since these lenses demand that the \emph{get} and \emph{put} functions form a bijection, 
no information is lost during translation from one data format to another. 
While this very strong guarantee can be useful, more often
some of the information in one or both of the data formats is
superfluous. For instance, the number of white space characters
between data items, the capitalization of various strings, the
sequence of fields in record, or the order of items in a list are
often irrelevant.  Many
data formats have their own ad hoc rules as well. In \bibtex{}, for example,
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---
the specific choice does not matter.

{\em Quotient lenses}~\cite{quotientlenses} loosen the restrictions of
bijective string lenses by including equivalence
relations on the source and target formats so that the round-tripping lens laws
hold modulo these equivalence rather than modulo equality. Quotient lenses thus
make it possible to synchronize many more data formats than would otherwise be
possible by ignoring details deemed inessential.

The bidirectional programming language {\em
Boomerang}~\cite{boomerang} is the only language we know of that
implements a general class of quotient lenses.  Unfortunately, these
lenses to have two main drawbacks.

The first drawback is that defining an equivalence
relation in Boomerang requires specifying a {\em canonizer}, which is
a pair of functions $f$ and $g$ where $f$
normalizes data from the
external format by choosing canonical representations
and $g$ provides an
inverse by specifying the external form of each canonical
representation. 
The disadvantage of this approach is that writing
a quotient lens in Boomerang requires defining four
different data formats: two ``external'' formats which specify the data to
be transformed by the lens, and two ``internal'' formats which specify the
canonical representations for the source and target data.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.

The second drawback with writing quotient lenses in Boomerang is that
even after specifying the four required data formats, the programmer must
still define a lens for converting between the canonical
representations.   Experience shows that writing such lenses can be
tedious 
(consider having to rearrange the order of data items by recursively
using operators to \swap adjacent fields) 
and that satisfying the type checker can be difficult because 
many seemingly useful lenses contain ambiguities disallowed by the 
Boomerang type checker~\cite{optician}.

Our solution to this problem is to introduce {\em QRE lenses}, which
are a new class of quotient lens that can be automatically
synthesized, obviating the need to write lenses by hand.  A QRE lens
is like a normal lens except it maps between formats described by
QREs. We prove that every composition of QRE lenses can be rewritten
to be the composition of a source canonizer, a bijective lens, and a
target canonizer. This normalization property means we can synthesize
QRE lenses by extending an existing algorithm for synthesizing
bijective lenses~\cite{optician}.

Generating a QRE lens requires only a pair of QREs to describe the
source and target formats and a (possibly empty) suite of examples
demonstrating the mapping.  For example, the following code

\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
\end{lstlisting}

binds \codefont{l} to a synthesized QRE lens mapping between names in
the comma-separated form describd by the QRE \codefont{comma_name}
and the space-separated form described by the
QRE \codefont{space_name}.

Given this information, we can exploit and
extend recent work on algorithms for synthesis of bijective
lenses~\cite{optician} to generate quotient lenses. The lenses generated by
our synthesis algorithm are guaranteed to translate data back and forth
between source and target QREs and to satisfy the provided examples.
Moreover, while our synthesized quotient lenses have a relatively
restricted form (which cuts down the amount of non-determinism in
a specification and greatly speeds up synthesis), we prove that
this rigidity does not compromise their expressiveness.

@@

\section{QRE Lenses by Example}
\label{sec:example}
\subsection{Bijective String Lenses}
Consider a bidirectional transformation that converts \bibtex{} citation records
such as
\begin{verbatim}
@Book {Lovelace,
Author = "Ada Lovelace",
Title = {Generic Title}
}
\end{verbatim}
\noindent
into an equivalent EndNote format like the following
\begin{verbatim}
%0 Book
%F Lovelace
%A Ada Lovelace
%T Generic Title
\end{verbatim}
\noindent
and vice versa.

@@

\begin{center}
\begin{minipage}{2.5in}
\begin{lstlisting}
bib_example =
"@Book {Lovelace,
 Author = \"Ada Lovelace\",
 Title = {Generic Title},
}"
\end{lstlisting}
\end{minipage}
\begin{minipage}{1.5in}
\begin{lstlisting}
end_example =
"%0 Book
 %T Generic Title
 %A Ada Lovelace
 %F Lovelace"
\end{lstlisting}
\end{minipage}
\end{center}
